Relative parabolic factorizations of Chevalley groups

### 2015-05-06 bass-kolster.tex
Minor fixes

\noindent Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=  \G(\Delta_1,R,I)\cdot \U(\Sigma_1^+,I)\cdot \U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I)\cdot\Z \cdot \U(\Phi^-,I)\cdot \U(\Phi^+,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \right\}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$. It is clear that vector $v=g\cdot v^+$ is $I$-unimodular.
By relative Chevalley---Matsumoto lemma it suffices to show that the set $\Z \cdot \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ contains an element $h$ such that $\mu_1(h\cdot v^+)=1$, $\mu_{\varpi\ssub{1} - \alpha\ssub{max}}(h \cdot v^+)=0$.
We find such $h$ by analyzing several cases.

>>>>

\noindent Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=  \G(\Delta_1,R,I)\cdot \U(\Sigma_1^+,I)\cdot \U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I)\cdot\Z \cdot \U(\Phi^-,I)\cdot \U(\Phi^+,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \right\}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$. It is clear that vector $v=g\cdot v^+$ is $I$-unimodular.
First of all, notice that it suffices to find $g' \in \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ such that $(g'\cdot v^+)_{1} = 1 + (g'\cdot v^+)_{\varpi\ssub{1}-\alpha\ssub{max}}$.
Indeed, multiplying 
By relative Chevalley---Matsumoto lemma it suffices to show that the set $\Z \cdot  \cdot g$ contains an element $h$ such that
$\mu_1(h\cdot v^+)=1$, $\mu_{\varpi\ssub{1} - \alpha\ssub{max}}(h \cdot v^+)=0$.
We find such $h$ by analyzing several cases.

### 2015-05-06 paper.tex
Minor fixes

It will be convenient to use standard numbering for the weights of natural representations of classical groups (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots l, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

For a weight $\lambda\in \Lambda(\pi)$ we denote by $\mu_{\lambda}\colon V\rightarrow R$ the coordinate function corresponding $\lambda$.
In other words, $\mu_\lambda(v)$ is the coefficient of the basis vector $e_\lambda$ in the expansion of $v\in V$ in $E_\pi$.
We denote by $v^+$ the basis vector corresponding to highest weight. Using the above notation for classical $\Phi$ we have $v^+=e_1$.
%TODO: State Matsumoto lemma and Chevalley-Matsumoto lemma

\subsection{Elementary subgroup}
Recall that elementary root unipotents $t_{\alpha}(\xi)$, $x\in R$, $\alpha\in\Phi$ satisfy the following set of relations called Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

@@

The following theorem is one of the main results of \cite{S} (see~\cite[Theorem~3.4]{S}).
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system and let $S$ be any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the following two families of elements:
\begin{itemize}
 \item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}

>>>>

It will be convenient to use standard numbering for the weights of natural representations of classical groups (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, \ell+1 & \text{in the case} & \Phi =\rA_\ell, \\
  1,2,\ldots \ell, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots \ell, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

For a weight $\lambda\in \Lambda(\pi)$ we denote by $(-)_\lambda\colon V\rightarrow R$ the coordinate function corresponding to $\lambda$.
In other words,  $v_\lambda$ is the coefficient of the basis vector $e_\lambda$ in the expansion of $v$ in $E_\pi$. 
Obviously, $v=\sum_{\lambda\in \Lambda(\pi)} v_\lambda e_\lambda$.
We denote by $v^+$ the basis vector corresponding to highest weight.
For classical $\Phi$ we have $v^+=e_1$.
%TODO: State Matsumoto lemma and Chevalley-Matsumoto lemma

\subsection{Elementary subgroup}
Recall from~\cite{St78, VP} that for $\alpha\in \Phi$, $\xi\in R$ one can define certain elements $t_{\alpha}(\xi)$ of $\G_sc(\Phi, R)$ called {\it elementary root unipotents}.
These elements satisfy well-known Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

@@

Let $I\leq R$ be an ideal. Denote by $\E(\Phi, I)$ the subgroup of $\G_{sc}(\Phi, R)$ generated by all elements $t_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$.
In the case $I=R$ the subgroup $\E(\Phi, R)$ is called the {\it elementary subgroup} of the Chevalley group $\G_{sc}(\Phi, R)$.

Denote by $z_\alpha(s, \xi)$ the element $x_\alpha(s)^{t\ssub{-\alpha}(\xi)}$.

\begin{thm}[L.~Vaserstein, J.~Tits] The relative elementary subgroup $\E(\Phi, R, I)$ is generated by elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$.
\end{thm}

In fact, one can choose a more economical set of generators for the elementary subgroup (see~\cite[Theorem~3.4]{S}).
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system and let $S$ be any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the following two families of elements:
\begin{itemize}
 \item $t_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}
